---
title: "P8106 Midterm Project"
author: "Chenxi Liu, Haoyi Tan,Haochen Yu"
date: "2026-03-29"
output:
  html_document:
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Overview

This analysis treats `log_PFS` as a continuous outcome and compares several candidate prediction models on the 1990--1995 cohort (`datA`). Model development and tuning are performed entirely within `datA` using repeated 10-fold cross-validation. The final model selected from `datA` is then externally validated on the 2000--2005 cohort (`datB`). If the selected model does not generalize well, the same candidate-model framework is refit in `datB`.

# 2. Data Loading and Preprocessing

```{r data_prep}
library(glmnet)
library(caret)
library(splines)
library(ggplot2)
library(corrplot)
library(earth)
library(mgcv)
library(gam)
library(mgcv)
library(dplyr)
library(tidyr)
library(gridExtra)

set.seed(2026)

# Load datasets
load("datA.RData")
load("datB.RData")

# Function to convert categorical variables to factors
convert_factors <- function(df) {
  df$menopause <- factor(df$menopause, levels = c(0,1), labels = c("Pre","Post"))
  df$race <- factor(df$race, levels = 1:4, labels = c("White","Black","Asian","Hispanic"))
  df$ER <- factor(df$ER, levels = c(0,1), labels = c("Negative","Positive"))
  df$PR <- factor(df$PR, levels = c(0,1), labels = c("Negative","Positive"))
  df$HER2 <- factor(df$HER2, levels = c(0,1), labels = c("Negative","Positive"))
  df$tumor_grade <- factor(df$tumor_grade, levels = 1:3, labels = c("Well","Moderate","Poor"))
  df$stage <- factor(df$stage, levels = 1:3, labels = c("I","II","III"))
  df$surgery <- factor(df$surgery, levels = c(0,1), labels = c("BCS","Mastectomy"))
  df$chemo <- factor(df$chemo, levels = c(0,1), labels = c("No","Yes"))
  df$hormonal_therapy <- factor(df$hormonal_therapy, levels = c(0,1), labels = c("No","Yes"))
  return(df)
}

datA_f <- convert_factors(datA)
datB_f <- convert_factors(datB)

# Remove ID column as it is not a predictor
datA_f$id <- NULL
datB_f$id <- NULL

# Check missingness, then use complete-case analysis if needed
cat("=== Missing values in datA ===\n")
print(colSums(is.na(datA_f)))
cat("\n=== Missing values in datB ===\n")
print(colSums(is.na(datB_f)))

nA_before <- nrow(datA_f)
nB_before <- nrow(datB_f)
datA_f <- na.omit(datA_f)
datB_f <- na.omit(datB_f)
cat("\ndatA rows before/after na.omit:", nA_before, nrow(datA_f), "\n")
cat("datB rows before/after na.omit:", nB_before, nrow(datB_f), "\n")

# Summary statistics
cat("\n=== datA Summary ===\n")
str(datA_f)
summary(datA_f$log_PFS)

cat("\n=== datB Summary ===\n")
str(datB_f)
summary(datB_f$log_PFS)
```

# 3. Exploratory Data Analysis

```{r eda, fig.width=8, fig.height=5}
# Distribution of log_PFS
par(mfrow = c(1, 2))
hist(datA_f$log_PFS, breaks = 30, main = "Distribution of log(PFS) - datA",
     xlab = "log(PFS)", col = "lightblue", border = "white")
hist(datB_f$log_PFS, breaks = 30, main = "Distribution of log(PFS) - datB",
     xlab = "log(PFS)", col = "lightgreen", border = "white")
par(mfrow = c(1, 1))

# Correlation plot for numeric variables
numeric_vars <- datA_f[, c("age", "bmi", "tumor_size", "lymph_nodes", "Ki67", "log_PFS")]
corrplot(cor(numeric_vars), method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix (datA)", mar = c(0,0,1,0))
```

# 4. Candidate Models and Tuning Grids

The following models are considered:

- OLS (baseline)
- Linear regression with natural cubic splines
- GAM
- Ridge regression
- Lasso regression
- Elastic net
- PCR
- PLS
- MARS

RMSE is used as the primary model-selection criterion in `datA`. MAE and R-squared are also reported for external validation in `datB`.

```{r tuning_grids}
# Set up cross-validation: 10-fold CV repeated 5 times
ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 5)

# Prepare model matrix
x_A <- model.matrix(log_PFS ~ ., data = datA_f)[, -1]
y_A <- datA_f$log_PFS
x_B_raw <- model.matrix(log_PFS ~ ., data = datB_f)[, -1]
y_B <- datB_f$log_PFS

# Align datB model matrix columns to datA model matrix columns
missing_cols <- setdiff(colnames(x_A), colnames(x_B_raw))
if (length(missing_cols) > 0) {
  x_B_raw <- cbind(x_B_raw,
                   matrix(0, nrow = nrow(x_B_raw), ncol = length(missing_cols),
                          dimnames = list(NULL, missing_cols)))
}
extra_cols <- setdiff(colnames(x_B_raw), colnames(x_A))
if (length(extra_cols) > 0) {
  x_B_raw <- x_B_raw[, setdiff(colnames(x_B_raw), extra_cols), drop = FALSE]
}
x_B <- x_B_raw[, colnames(x_A), drop = FALSE]

# Unified tuning grids for penalized models
lambda_grid <- 10^seq(-3, 1, length = 100)
ridge_grid <- expand.grid(alpha = 0, lambda = lambda_grid)
lasso_grid <- expand.grid(alpha = 1, lambda = lambda_grid)
enet_grid  <- expand.grid(alpha = c(0.25, 0.50, 0.75), lambda = lambda_grid)
mars_grid  <- expand.grid(degree = 1:2, nprune = seq(5, 25, by = 5))

# Formulas for interpretable models
formula_ols <- log_PFS ~ .
formula_spline <- log_PFS ~ 
  ns(age, df = 3) + menopause + race + ns(bmi, df = 3) +
  ns(tumor_size, df = 3) + tumor_grade + ns(lymph_nodes, df = 3) +
  ER + PR + HER2 + ns(Ki67, df = 3) + stage + surgery + chemo +
  hormonal_therapy

# GAM allows flexible nonlinear effects for continuous predictors
formula_gam <- log_PFS ~ 
  s(age) + menopause + race + s(bmi) + s(tumor_size) + tumor_grade +
  s(lymph_nodes) + ER + PR + HER2 + s(Ki67) + stage + surgery +
  chemo + hormonal_therapy

tuning_table <- data.frame(
  Model = c("OLS", "Spline LM", "GAM", "Ridge", "Lasso", "Elastic Net", "PCR", "PLS", "MARS"),
  Tuning_Parameters = c(
    "None",
    "Natural spline df fixed at 3 for each continuous predictor",
    "Smoothing parameters estimated automatically by the fitting procedure",
    "lambda",
    "lambda",
    "alpha and lambda",
    "ncomp",
    "ncomp",
    "degree and nprune"
  ),
  Grid = c(
    "None",
    "df = 3 (prespecified)",
    "Automatic smoothing selection",
    "lambda = 10^seq(-3, 1, length = 100)",
    "lambda = 10^seq(-3, 1, length = 100)",
    "alpha = {0.25, 0.50, 0.75}; lambda = 10^seq(-3, 1, length = 100)",
    paste0("ncomp = 1:", ncol(x_A)),
    paste0("ncomp = 1:", min(20, ncol(x_A))),
    "degree = {1,2}; nprune = {5,10,15,20,25}"
  )
)
knitr::kable(tuning_table, caption = "Candidate models and tuning grids")
```

# 5. Part I: Model Training on 1990-1995 Cohort (datA)

```{r model_training, cache=TRUE}
# 1. OLS
# OLS remains the baseline linear model
ols_fit <- train(formula_ols, data = datA_f, method = "lm", trControl = ctrl)

# 2. Linear regression with natural cubic splines
# Splines allow flexible but still regression-based nonlinear effects
spline_fit <- train(formula_spline, data = datA_f, method = "lm", trControl = ctrl)

# 3. GAM
# GAM allows flexible nonlinear effects for continuous predictors
# Use mgcv::gam directly because caret::train(method = "gam") can be unstable across environments
gam_fit <- mgcv::gam(formula_gam, data = datA_f, method = "REML")

# Manual repeated 10-fold CV for GAM to make its evaluation comparable to other models
set.seed(123)
gam_folds_A <- caret::createMultiFolds(y_A, k = 10, times = 5)
gam_pred_A <- vector("list", length(gam_folds_A))
gam_obs_A  <- vector("list", length(gam_folds_A))

for (i in seq_along(gam_folds_A)) {
  train_idx <- gam_folds_A[[i]]
  test_idx  <- setdiff(seq_len(nrow(datA_f)), train_idx)
  fit_i <- mgcv::gam(formula_gam, data = datA_f[train_idx, , drop = FALSE], method = "REML")
  pred_i <- predict(fit_i, newdata = datA_f[test_idx, , drop = FALSE])
  gam_pred_A[[i]] <- pred_i
  gam_obs_A[[i]]  <- datA_f$log_PFS[test_idx]
}

gam_pred_vec_A <- unlist(gam_pred_A)
gam_obs_vec_A  <- unlist(gam_obs_A)
gam_results <- data.frame(
  RMSE = sqrt(mean((gam_obs_vec_A - gam_pred_vec_A)^2)),
  Rsquared = 1 - sum((gam_obs_vec_A - gam_pred_vec_A)^2) / sum((gam_obs_vec_A - mean(gam_obs_vec_A))^2),
  MAE = mean(abs(gam_obs_vec_A - gam_pred_vec_A))
)
cat("GAM CV RMSE:", gam_results$RMSE, "\n")

# 4. Ridge Regression
ridge_fit <- train(x = x_A, y = y_A, method = "glmnet",
                   tuneGrid = ridge_grid, trControl = ctrl,
                   standardize = TRUE)
cat("Best Ridge lambda:", ridge_fit$bestTune$lambda, "\n")

# 5. Lasso Regression
lasso_fit <- train(x = x_A, y = y_A, method = "glmnet",
                   tuneGrid = lasso_grid, trControl = ctrl,
                   standardize = TRUE)
cat("Best Lasso lambda:", lasso_fit$bestTune$lambda, "\n")

# 6. Elastic Net
enet_fit <- train(x = x_A, y = y_A, method = "glmnet",
                  tuneGrid = enet_grid, trControl = ctrl,
                  standardize = TRUE)
cat("Best Elastic Net alpha:", enet_fit$bestTune$alpha, 
    "lambda:", enet_fit$bestTune$lambda, "\n")

# 7. PCR
pcr_fit <- train(x = x_A, y = y_A, method = "pcr",
                 tuneGrid = data.frame(ncomp = 1:ncol(x_A)),
                 trControl = ctrl, preProcess = c("center", "scale"))
cat("Best PCR ncomp:", pcr_fit$bestTune$ncomp, "\n")

# 8. PLS
pls_fit <- train(x = x_A, y = y_A, method = "pls",
                 tuneGrid = data.frame(ncomp = 1:min(20, ncol(x_A))),
                 trControl = ctrl, preProcess = c("center", "scale"))
cat("Best PLS ncomp:", pls_fit$bestTune$ncomp, "\n")

# 9. MARS
mars_fit <- train(x = x_A, y = y_A, method = "earth",
                  tuneGrid = mars_grid, trControl = ctrl)
cat("Best MARS degree:", mars_fit$bestTune$degree, 
    "nprune:", mars_fit$bestTune$nprune, "\n")
```

# 6. Model Comparison and Selection

```{r model_comparison, fig.width=10, fig.height=6}
# Resamples comparison
resamp <- resamples(list(
  OLS = ols_fit,
  Spline = spline_fit,
  Ridge = ridge_fit,
  Lasso = lasso_fit,
  ElasticNet = enet_fit,
  PCR = pcr_fit,
  PLS = pls_fit,
  MARS = mars_fit
))

# Summary
resamp_summary <- summary(resamp)
resamp_summary

# Figure 1: CV RMSE boxplot comparison
bwplot(resamp, metric = "RMSE", main = "Cross-Validation RMSE Comparison (caret models)")
cat("\nNote: GAM was fit with mgcv::gam and evaluated by manual repeated 10-fold CV; its RMSE is reported in the summary table below.\n")
```

```{r final_model_selection}
# Collect the best resampling result for each model
cv_results <- bind_rows(
  ols_fit$results %>% mutate(Model = "OLS"),
  spline_fit$results %>% mutate(Model = "Spline"),
  gam_results %>% mutate(Model = "GAM"),
  ridge_fit$results %>% filter(alpha == ridge_fit$bestTune$alpha,
                               lambda == ridge_fit$bestTune$lambda) %>% mutate(Model = "Ridge"),
  lasso_fit$results %>% filter(alpha == lasso_fit$bestTune$alpha,
                               lambda == lasso_fit$bestTune$lambda) %>% mutate(Model = "Lasso"),
  enet_fit$results %>% filter(alpha == enet_fit$bestTune$alpha,
                              lambda == enet_fit$bestTune$lambda) %>% mutate(Model = "ElasticNet"),
  pcr_fit$results %>% filter(ncomp == pcr_fit$bestTune$ncomp) %>% mutate(Model = "PCR"),
  pls_fit$results %>% filter(ncomp == pls_fit$bestTune$ncomp) %>% mutate(Model = "PLS"),
  mars_fit$results %>% filter(degree == mars_fit$bestTune$degree,
                              nprune == mars_fit$bestTune$nprune) %>% mutate(Model = "MARS")
) %>%
  select(Model, RMSE, Rsquared, MAE, everything()) %>%
  arrange(RMSE)

knitr::kable(cv_results[, c("Model", "RMSE", "Rsquared", "MAE")],
             digits = 4,
             caption = "Cross-validated performance on datA (ranked by RMSE)")

best_model_name <- cv_results$Model[1]
cat("Final model selected from datA:", best_model_name, "\n")

# Model list for downstream prediction
models <- list(
  OLS = ols_fit,
  Spline = spline_fit,
  GAM = gam_fit,
  Ridge = ridge_fit,
  Lasso = lasso_fit,
  ElasticNet = enet_fit,
  PCR = pcr_fit,
  PLS = pls_fit,
  MARS = mars_fit
)
final_model <- models[[best_model_name]]
```

```{r interpretation_outputs, fig.width=10, fig.height=4}
# Baseline OLS model for interpretation
ols_interpret <- lm(formula_ols, data = datA_f)
summary(ols_interpret)

# OLS diagnostic plots
par(mfrow = c(1, 2))
plot(ols_interpret, which = 1)
plot(ols_interpret, which = 2)
par(mfrow = c(1, 1))

# Lasso coefficient path and coefficients (useful for interpretation if penalized methods perform well)
cat("\n=== Lasso Coefficients ===\n")
lasso_coef <- coef(lasso_fit$finalModel, lasso_fit$bestTune$lambda)
print(lasso_coef)

plot(lasso_fit$finalModel, xvar = "lambda", label = TRUE,
     main = "Lasso Coefficient Path")
```

# 7. Part II: Generalization to 2000-2005 Cohort (datB)

```{r generalization}
# Function to calculate metrics
calc_metrics <- function(actual, predicted) {
  rmse <- sqrt(mean((actual - predicted)^2))
  mae <- mean(abs(actual - predicted))
  rsq <- 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
  return(c(RMSE = rmse, MAE = mae, Rsquared = rsq))
}

# Prediction helper: formula-based models use datB_f, matrix-based models use x_B
predict_from_model <- function(name, model_object, datB_f, x_B) {
  if (name %in% c("OLS", "Spline", "GAM")) {
    predict(model_object, newdata = datB_f)
  } else {
    predict(model_object, newdata = x_B)
  }
}

# External validation of the final model selected from datA
pred_B_final <- predict_from_model(best_model_name, final_model, datB_f, x_B)
final_metrics_B <- calc_metrics(y_B, pred_B_final)

final_validation <- data.frame(
  Model = best_model_name,
  RMSE = final_metrics_B["RMSE"],
  MAE = final_metrics_B["MAE"],
  Rsquared = final_metrics_B["Rsquared"]
)
cat("=== External validation of final datA model on datB ===\n")
print(final_validation, row.names = FALSE)

# Calibration assessment
calibration_fit <- lm(y_B ~ pred_B_final)
cat("\n=== Calibration model for final datA model on datB ===\n")
summary(calibration_fit)
```

```{r generalization_plots, fig.width=10, fig.height=5}
# Optional: external validation results for all datA models
all_external_results <- data.frame()
for (name in names(models)) {
  pred <- predict_from_model(name, models[[name]], datB_f, x_B)
  metrics <- calc_metrics(y_B, pred)
  all_external_results <- rbind(all_external_results,
                                data.frame(Model = name,
                                           RMSE = metrics["RMSE"],
                                           MAE = metrics["MAE"],
                                           Rsquared = metrics["Rsquared"]))
}
all_external_results <- all_external_results[order(all_external_results$RMSE), ]
cat("\n=== External validation results for all datA models on datB ===\n")
print(all_external_results, row.names = FALSE)

# Figure 3: Predicted vs Actual for final model
par(mfrow = c(1, 2))
plot(pred_B_final, y_B, pch = 20, col = rgb(0, 0, 1, 0.3),
     xlab = "Predicted log(PFS)", ylab = "Observed log(PFS)",
     main = paste0(best_model_name, ": Predicted vs Observed (datB)"))
abline(0, 1, col = "red", lwd = 2)

plot(pred_B_final, y_B - pred_B_final, pch = 20, col = rgb(0, 0.5, 0, 0.3),
     xlab = "Predicted log(PFS)", ylab = "Residuals",
     main = paste0(best_model_name, ": Residuals (datB)"))
abline(h = 0, col = "red", lwd = 2)
par(mfrow = c(1, 1))
```

# 8. Building New Model on datB

```{r new_model_datB, cache=TRUE}
ctrl_B <- trainControl(method = "repeatedcv", number = 10, repeats = 5)

# Refit the same candidate-model set on datB
ols_fit_B <- train(formula_ols, data = datB_f, method = "lm", trControl = ctrl_B)
spline_fit_B <- train(formula_spline, data = datB_f, method = "lm", trControl = ctrl_B)
# Refit GAM on datB using mgcv::gam and evaluate it by manual repeated 10-fold CV
gam_fit_B <- mgcv::gam(formula_gam, data = datB_f, method = "REML")

set.seed(123)
gam_folds_B <- caret::createMultiFolds(y_B, k = 10, times = 5)
gam_pred_B_cv <- vector("list", length(gam_folds_B))
gam_obs_B_cv  <- vector("list", length(gam_folds_B))

for (i in seq_along(gam_folds_B)) {
  train_idx <- gam_folds_B[[i]]
  test_idx  <- setdiff(seq_len(nrow(datB_f)), train_idx)
  fit_i <- mgcv::gam(formula_gam, data = datB_f[train_idx, , drop = FALSE], method = "REML")
  pred_i <- predict(fit_i, newdata = datB_f[test_idx, , drop = FALSE])
  gam_pred_B_cv[[i]] <- pred_i
  gam_obs_B_cv[[i]]  <- datB_f$log_PFS[test_idx]
}

gam_pred_vec_B <- unlist(gam_pred_B_cv)
gam_obs_vec_B  <- unlist(gam_obs_B_cv)
gam_results_B <- data.frame(
  RMSE = sqrt(mean((gam_obs_vec_B - gam_pred_vec_B)^2)),
  Rsquared = 1 - sum((gam_obs_vec_B - gam_pred_vec_B)^2) / sum((gam_obs_vec_B - mean(gam_obs_vec_B))^2),
  MAE = mean(abs(gam_obs_vec_B - gam_pred_vec_B))
)

ridge_fit_B <- train(x = x_B, y = y_B, method = "glmnet",
                     tuneGrid = ridge_grid, trControl = ctrl_B,
                     standardize = TRUE)
lasso_fit_B <- train(x = x_B, y = y_B, method = "glmnet",
                     tuneGrid = lasso_grid, trControl = ctrl_B,
                     standardize = TRUE)
enet_fit_B <- train(x = x_B, y = y_B, method = "glmnet",
                    tuneGrid = enet_grid, trControl = ctrl_B,
                    standardize = TRUE)
pcr_fit_B <- train(x = x_B, y = y_B, method = "pcr",
                   tuneGrid = data.frame(ncomp = 1:ncol(x_B)),
                   trControl = ctrl_B, preProcess = c("center", "scale"))
pls_fit_B <- train(x = x_B, y = y_B, method = "pls",
                   tuneGrid = data.frame(ncomp = 1:min(20, ncol(x_B))),
                   trControl = ctrl_B, preProcess = c("center", "scale"))
mars_fit_B <- train(x = x_B, y = y_B, method = "earth",
                    tuneGrid = mars_grid, trControl = ctrl_B)
```

```{r new_model_datB_compare, fig.width=10, fig.height=6}
cv_results_B <- bind_rows(
  ols_fit_B$results %>% mutate(Model = "OLS"),
  spline_fit_B$results %>% mutate(Model = "Spline"),
  gam_results_B %>% mutate(Model = "GAM"),
  ridge_fit_B$results %>% filter(alpha == ridge_fit_B$bestTune$alpha,
                                 lambda == ridge_fit_B$bestTune$lambda) %>% mutate(Model = "Ridge"),
  lasso_fit_B$results %>% filter(alpha == lasso_fit_B$bestTune$alpha,
                                 lambda == lasso_fit_B$bestTune$lambda) %>% mutate(Model = "Lasso"),
  enet_fit_B$results %>% filter(alpha == enet_fit_B$bestTune$alpha,
                                lambda == enet_fit_B$bestTune$lambda) %>% mutate(Model = "ElasticNet"),
  pcr_fit_B$results %>% filter(ncomp == pcr_fit_B$bestTune$ncomp) %>% mutate(Model = "PCR"),
  pls_fit_B$results %>% filter(ncomp == pls_fit_B$bestTune$ncomp) %>% mutate(Model = "PLS"),
  mars_fit_B$results %>% filter(degree == mars_fit_B$bestTune$degree,
                                nprune == mars_fit_B$bestTune$nprune) %>% mutate(Model = "MARS")
) %>%
  select(Model, RMSE, Rsquared, MAE, everything()) %>%
  arrange(RMSE)

knitr::kable(cv_results_B[, c("Model", "RMSE", "Rsquared", "MAE")],
             digits = 4,
             caption = "Cross-validated performance on datB after refitting")

best_model_name_B <- cv_results_B$Model[1]
cat("Best refit model on datB:", best_model_name_B, "\n")
```

```{r coefficient_comparison, fig.width=10, fig.height=7}
# Coefficient comparison for lasso as an example of how selected effects may change over time
coef_A_lasso <- as.matrix(coef(lasso_fit$finalModel, s = lasso_fit$bestTune$lambda))
coef_B_lasso <- as.matrix(coef(lasso_fit_B$finalModel, s = lasso_fit_B$bestTune$lambda))
common_vars <- intersect(rownames(coef_A_lasso), rownames(coef_B_lasso))
coef_compare <- data.frame(
  Variable = common_vars[common_vars != "(Intercept)"],
  datA = coef_A_lasso[common_vars[common_vars != "(Intercept)"], 1],
  datB = coef_B_lasso[common_vars[common_vars != "(Intercept)"], 1]
)

coef_long <- pivot_longer(coef_compare,
                          cols = c(datA, datB),
                          names_to = "Dataset",
                          values_to = "Coefficient")

ggplot(coef_long, aes(x = Coefficient, y = reorder(Variable, abs(Coefficient)), fill = Dataset)) +
  geom_col(position = "dodge") +
  labs(title = "Lasso coefficients: datA versus datB",
       x = "Coefficient", y = "Variable")
```
